---
title: "test_tree_thinning"
author: "Janet Young"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, results=FALSE, warning=FALSE, message=FALSE, error=FALSE}
library(tidyverse)
library(phytools)    ## for the mammal.tree dataset
library(ggtree)
library(tidytree)
library(dendextend)  ## for a cutree function that works on mammal.tree
```

Load a test tree and show some basic information on it. It's an object of class `phylo`.

```{r load data}
data(mammal.tree) #  load example tree from phytools

mammal.tree
```

Plot the tree (plot function comes from the ape package, I think - see ?plot.phylo). 

```{r simple tree plot, fig.height=5, width=5}
plot(mammal.tree, cex=0.5)
add.scale.bar()
```


We'll use the `cutree` function to slice the tree into some number of clusters (5 in this example): it returns a vector of cluster IDs for each taxon, but I turn that into a tibble.

```{r define clades}
cladeIDs <- dendextend::cutree(mammal.tree, k=5) %>% 
    as_tibble(rownames="taxon") %>% 
    dplyr::rename(clusterID=value) %>% 
    mutate(clusterID=factor(paste("cluster_",clusterID,sep="")))

cladeIDs %>% 
    head()
```
Tabulate cluster membership
```{r tabulate cluster membership}
cladeIDs %>% 
    count(clusterID)
```

`ggtree` is a nice package for tree plotting. The [ggtree 'book' website](https://yulab-smu.top/treedata-book/index.html) is helpful.

Here's that simple plot again using ggtree:
```{r siple ggtree plot, fig.height=5, width=5}
ggtree(mammal.tree) +
    geom_tiplab(size=2.5) +
    hexpand(0.5)
```

But ggtree has lots of cool ways to display more info about the tree.

Let's say we have info on the taxa in a tree, and maybe we've stored it in a tibble. To use ggtree's plotting functions, we want to make a single object that contains the tree and the tibble together - that will be an object of class `treedata` (defined by the `tidytree` package)

Maybe there's an easier way to do that, but I wrote a function called `addInfoToTree` to merge tree and taxon info that includes a couple of sanity checks.  `left_join` is the function that actually merges the tree and the information. 

```{r define addInfoToTree function}
###### addInfoToTree - make a treedata object by combining a phylo object with a tibble that has info on the tips
addInfoToTree <- function(tree, info, colnameForTaxonLabels="taxon") {
    ##### get info in same order as taxa in the tree:
    if (! colnameForTaxonLabels %in% colnames(info)) {
        stop("\n\nERROR - there should be a ",colnameForTaxonLabels," column in the info table\n\n")
    }
    ## check all taxa are in the info table
    tipLabelsInInfoTable <- info %>% select(all_of(colnameForTaxonLabels)) %>% deframe()
    if(length(setdiff(tree$tip.label, tipLabelsInInfoTable))>0) {
        stop("\n\nERROR - there are taxon labels in the tree that are not in the info table\n\n")
    }
    # now get info
    desiredRows <- match(tree$tip.label, tipLabelsInInfoTable)
    info_treeorder <- info[desiredRows,] 
    # add info to tree
    tree_withInfo <-  left_join(
        tree, 
        info_treeorder ,
        by=c("label"=colnameForTaxonLabels))
    return(tree_withInfo)
}

```

Combine the mammal.tree with the cladeID tibble we made using cutree

```{r use addInfoToTree}
mammal.tree.withInfo <- addInfoToTree(mammal.tree, cladeIDs)
```

Plot the tree again and color the clades
```{r plot tree with colored clades, fig.height=5, width=5}
ggtree(mammal.tree.withInfo,aes(color=clusterID) ) +
    geom_tiplab(aes(color=clusterID), size=2.5) + 
    hexpand(0.5) +
    theme(legend.position = "top")
```


We could randomly choose one taxon per clade like this:
```{r choose one taxon per clade}
cladeMembersList <- split( cladeIDs$taxon, cladeIDs$clusterID)
oneMemberPerClade <- sapply(cladeMembersList, sample, size=1)
```

And we can update the info table associated with the tree to show which taxa we chose, and label those tips on the tree plot:
```{r label chosen taxa, fig.height=5, width=5}
mammal.tree.withInfo <- mammal.tree.withInfo %>% 
    mutate( chosen= label %in% oneMemberPerClade) 

ggtree(mammal.tree.withInfo,aes(color=clusterID) ) +
    geom_tiplab(aes(color=clusterID), size=2.5)+ 
    geom_tippoint(aes(subset= chosen), size=3) +
    hexpand(0.5) +
    theme(legend.position = "top")
```