---
title: "how to combine GrangesList into single set of regions"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(GenomicRanges)
```

Task : take a list of regions (e.g. each list item has regions identified in a single sample) and make a single set of regions found in any sample

I found a much simpler solution to the thing we struggled with on Friday while reading the help page `?GRangesList`. 

We used that weird `reduce(Reduce())` syntax, but `unlist` is a simpler way to do it.  

The help page says:
```
unlist(x, recursive = TRUE, use.names = TRUE):
    Concatenates the elements of x into a single GRanges object.
```


Make example objects:
```{r}
# make single GRanges object
gr1 <- GRanges(seqnames=c("chr1","chr1","chr2","chr2"),
               ranges=IRanges(start=c(1,100,1,100), width=20),
               score1=30)
# make ordinary list
gr_list <- list(first=gr1, second=gr1, third=gr1)
# make GRangesList
gr_GRlist <- GRangesList(gr_list)
```


`unlist`'s default is that it adds names to each region (shown on the left) to show which of the original objects they came from:
```{r}
unlist(gr_GRlist)
```

Notice that this doesn't work on a regular list object:
```{r}
unlist(gr_list)
```


You probably still want to use the `reduce()` function to merge overlapping intervals that are present in more than one sample:
```{r}
reduce(unlist(gr_GRlist))
```

Exploring `unlist()` a bit more (notice I'm not using `reduce()` now):

If I know I don't need to keep track of the original objects, I use the option `use.names = FALSE`.  Having duplicate names for regions can sometimes cause trouble down the road
```{r}
unlist(gr_GRlist, use.names = FALSE)
```

Sometimes I do want to keep track of the original objects, and I find it works better to do make a metadata column to contain that information:
```{r}
combined_gr_GRlist <- unlist(gr_GRlist)
# put the names in a metadata column
mcols(combined_gr_GRlist)[,"whichOriginalGR"] <- names(combined_gr_GRlist)
# remove the names
names(combined_gr_GRlist) <- NULL
combined_gr_GRlist
```
