---
title: "how to rename a Granges column"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(GenomicRanges)
```


Make example GRanges object and turn it into a list and a GRangesList:
```{r}
# make single Granges object
gr1 <- GRanges(seqnames=c("chr1","chr1","chr2","chr2"),
               ranges=IRanges(start=c(1,100,1,100), width=20),
               score1=30, score2=50, score3=1)
# make ordinary list
gr_list <- list(first=gr1, second=gr1, third=gr1)
# make GRangesList
gr_GRlist <- GRangesList(gr_list)
```

Show colnames before we make changes:
```{r}
colnames(mcols(gr1))
```


Now explore ways to change a column name in a **single** GRanges object:

```{r}
# change colnames using a find-replace (gsub) 
# the ^ and $ mean that score1 must be the entire name, not a partial match:
colnames(mcols(gr1)) <- gsub("^score1$","myNewColname", colnames(mcols(gr1)) )

# changes colnames in an exact match way:
colnames(mcols(gr1))[ which(colnames(mcols(gr1)) =="score2") ] <- "myNewColname2"

# changes colnames using position:
colnames(mcols(gr1))[ 3 ] <- "myNewColname3"

# can also used variables:
newColnameIwant <- "myNewColname3a"
colnames(mcols(gr1))[ 3 ] <- newColnameIwant

# show colnames now: 
colnames(mcols(gr1))
```

We can use any of those methods for LIST objects - for an ordinary list we do it via `lapply`:
```{r}
gr_list <- lapply(gr_list, function(x){
    colnames(mcols(x))[3] <- newColnameIwant
    return(x)
})
gr_list
```


For a GRangesList we do it via `endoapply`   

`endoapply` is similar to `lapply`, but it makes sure the output object has the same class as the input object. if we had used `lapply` here, output would have class `list`. with `endoapply`, output has class `GRangesList`
```{r}
gr_GRlist <- endoapply(gr_GRlist, function(x){
    colnames(mcols(x))[3] <- newColnameIwant
    return(x)
})
gr_GRlist
```

Sometimes I want to use the name of each element in the list to add it to the column names, like this:
```{r}
# because I'm running "apply" on the NAMES (class=character) not the GRangesList I can't use endoapply, because it will want to return an object of class character. So I use lapply and turn it back into GRangesList afterwards
gr_GRlist_new <- lapply(names(gr_GRlist), function(x){
    this_gr <- gr_GRlist[[x]]
    colnames(mcols(this_gr))[1] <- paste(x,"_score",sep="")
    return(this_gr)
})
# need to add back the names
names(gr_GRlist_new) <- names(gr_GRlist)


# turn it back into GRangesList . turns out this tries to make the colnames the same across all list elements. might result in extra columns. If that's a problem, we can try to figure out a way to suppress that  
gr_GRlist_new2 <- GRangesList(gr_GRlist_new)

gr_GRlist_new2
```

